Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-05-17 article-models-ttinternal.tex
initiated the section on substitution algebras
formulated the groupoid conjecture

@@

\subsection{Extension-weakening algebras}
The notion of (extension-)weakening algebra is dependent on the notion of extension algebra.
Although it is strictly speaking not dependent on the notion of empty-algebra,
we shall only formulate a weakening algebras in the setting of an
extension-empty algebras. When one wants to have a weakening operation which
also acts on the level of contexts in an extension algebra without empty
context, extra work has to be done to introduce these separately.

An extension-weakening algebra in context $\Gamma$ will be an octuple
\begin{equation*}
(A,P,\epsilon_0,\epsilon_1,\phi_0,\phi_1,\omega_0,\omega_1)
\end{equation*}
where $(A,P,\epsilon_0,\epsilon_1,\phi_0,\phi_1)$ is an extension-empty algebra
in context $\Gamma$ and where
\begin{align*}
\jhom*
  {{{\Gamma}{A}}{P}}
  {\ctxwk{P}{P}}
  {\jcomp{}{\epsilon_0}{P}}
  {\omega_0}
  \\
\jfhom*
  {{{\Gamma}{A}}{P}}
  {\ctxwk{P}{P}}
  {\jcomp{}{\epsilon_0}{P}}
  {\omega_0}
  {\ctxwk{P}{\jcomp{}{\epsilon_0}{P}}}
  {\jcomp{}{\epsilon_0}{\jcomp{}{\epsilon_0}{P}}}
  {\omega_1}
\end{align*}
such that $(\omega_0,\omega_1)$ is both an extension homomorphism
\begin{equation*}
\begin{tikzcd}
( \ctxwk{P}{P},
  \ctxwk{P}{\jcomp{}{\epsilon_0}{P}},
  \ctxwk{P}{\epsilon_1},
  \ctxwk{P}{\jcomp{}{\epsilon_0}{\epsilon_1}})
\ar{r}{(\omega_0,\omega_1)}
& ( \jcomp{}{\epsilon_0}{P},
  \jcomp{}{\epsilon_0}{\jcomp{}{\epsilon_0}{P}},
  \jcomp{}{\epsilon_0}{\epsilon_1},
  \jcomp{}{\epsilon_0}{\jcomp{}{\epsilon_0}{\epsilon_1}})
\end{tikzcd}
\end{equation*}
and an empty homomorphism

@@

>>>>

It would be interesting to write out a weak version of pre-universes, internal
to Martin-L\"of type theory with the function extensionality principle. 
To do this, the empty context needs to
be replaced by a contractible type, extension by dependent pair types,
judgmental equalities of terms by identifications and judgmental equalities
of families by equivalences of types. We conjecture that it is possible to
carry this out (in particular to make sure that all the constructions 
type-check). Moreover, one could then extend the notion of `weak' pre-universes
with the requirement that every internal morphism is weakly anodyne. This
should give an internal theory of weak higher groupoids.

@@

\subsection{Term-algebras and term-operations}
Until now we had no need to consider term-algebras. Extension is an operation
letting contexts and families interact and the empty object is a term which
requires that contexts and families are compatible. However, to model the
theory of contexts, families and terms we do need to take terms into account.

\begin{defn}
A term-algebra in context $\Gamma$ is a septuple 
$(A,P,T,\epsilon_0,\epsilon_1,\phi_0,\phi_1)$ such that the quadruple
$(A,P,\epsilon_0,\epsilon_1)$ is an extension algebra in context $\Gamma$, 
the quadruple $(A,P,\phi_0,\phi_1)$ is an empty algebra in context $\Gamma$,
for which we additionally have the following judgmental equality:
\begin{align*}
\jfameq*{{\Gamma}{A}}{\subst{\phi_1}{T}}{\emptyf}.
\end{align*}
\end{defn}

\begin{thm}
Let $\mathcal{A}\defeq(A,P,T,\epsilon_0,\epsilon_1,\phi_0,\phi_1)$ be a term-algebra in context
$\Gamma$. Then the septuple
\begin{equation*}
(P,\jcomp{}{\epsilon_0}{P},\jcomp{}{\epsilon_0}{T},\epsilon_1,\jcomp{}{\epsilon_0}{\epsilon_1},\phi_1,\jcomp{}{\epsilon_0}{\phi_1})
\end{equation*}
is a term-algebra in context $\ctxext{\Gamma}{A}$, called the family term-algebra
of $\mathcal{A}$.
\end{thm}

\begin{thm}
Let $\mathcal{A}\defeq(A,P,T,\epsilon_0,\epsilon_1,\phi_0,\phi_1)$ be a term-algebra in context
$\Gamma$ and let $X$ be a family in context $\Gamma$. Then
\begin{equation*}
\ctxwk{X}{\mathcal{A}}\defeq(\ctxwk{X}{A},\ctxwk{X}{P},\ctxwk{X}{T},\ctxwk{X}{\epsilon_0},\ctxwk{X}{\epsilon_1},\ctxwk{X}{\phi_0},\ctxwk{X}{\phi_1})
\end{equation*}
is a term-algebra in context $\ctxext{\Gamma}{X}$.
\end{thm}

\begin{thm}
Let $\mathcal{A}\defeq(A,P,T,\epsilon_0,\epsilon_1,\phi_0,\phi_1)$ be a term-algebra in context
$\ctxext{\Gamma}{X}$ and let $\jterm{\Gamma}{X}{x}$. Then
\begin{equation*}
\subst{x}{\mathcal{A}}\defeq
(\subst{x}{A},\subst{x}{P},\subst{x}{T},\subst{x}{\epsilon_0},\subst{x}{\epsilon_1},
\subst{x}{\phi_0},\subst{x}{\phi_1})
\end{equation*}
is a term-algebra in context $\Gamma$.
\end{thm}

\begin{defn}
Let $\mathcal{A}$ and $\mathcal{B}$ be term-algebras in context $\Gamma$. 
A term-operation from $\mathcal{A}$ to $\mathcal{B}$ is a triple $(f_0,f_1,f_2)$
consisting of
\begin{align*}
\jhom*{\Gamma}{A}{B}{f_0}
  \\
\jfhom*{\Gamma}{A}{B}{f_0}{P}{Q}{f_1}
  \\
\jfhom*{\Gamma}{{A}{P}}{{B}{Q}}{\jvcomp{P}{f_0}{f_1}}{T}{U}{f_2}
\end{align*}
such that the pair $(f_0,f_1)$ is both an extension homomorphism from
$(A,P,\epsilon_0,\epsilon_1)$ to $(B,Q,\eta_0,\eta_1)$ in context $\Gamma$
and an empty homomorphism from $(A,P,\phi_0,\phi_1)$ to $(B,Q,\psi_0,\psi_1)$
in context $\Gamma$.
\end{defn}

\subsection{Weakening term-algebras}
Weakening term-algebras will be term-algebras with certain added structure.
Although strictly speaking one could formulate a notion of weakening algebra
which only depends on extension algebras and which omits both empty objects
and terms, we shall not do so here.

Let $(A,P,T,\epsilon_0,\epsilon_1,\phi_0,\phi_1)$ be a term-algebra in context
$\Gamma$.

An extension-weakening algebra in context $\Gamma$ will be an octuple
\begin{equation*}
(A,P,\epsilon_0,\epsilon_1,\phi_0,\phi_1,\omega_0,\omega_1)
\end{equation*}
where $(A,P,\epsilon_0,\epsilon_1,\phi_0,\phi_1)$ is an extension-empty algebra
in context $\Gamma$ and where
\begin{align*}
\jhom*
  {{{\Gamma}{A}}{P}}
  {\ctxwk{P}{P}}
  {\jcomp{}{\epsilon_0}{P}}
  {\omega_0}
  \\
\jfhom*
  {{{\Gamma}{A}}{P}}
  {\ctxwk{P}{P}}
  {\jcomp{}{\epsilon_0}{P}}
  {\omega_0}
  {\ctxwk{P}{\jcomp{}{\epsilon_0}{P}}}
  {\jcomp{}{\epsilon_0}{\jcomp{}{\epsilon_0}{P}}}
  {\omega_1}
\end{align*}
such that $(\omega_0,\omega_1)$ is both an extension homomorphism
\begin{equation*}
\begin{tikzcd}[column sep=large]
\ctxwk{P}{(P,
  \jcomp{}{\epsilon_0}{P},
  \epsilon_1,
  \jcomp{}{\epsilon_0}{\epsilon_1})}
\ar{r}{(\omega_0,\omega_1)}
& \jcomp{}{\epsilon_0}{(P,
  \jcomp{}{\epsilon_0}{P},
  \epsilon_1,
  \jcomp{}{\epsilon_0}{\epsilon_1})}
\end{tikzcd}
\end{equation*}
and an empty homomorphism

@@

Let $(A,P,T,\epsilon_0,\epsilon_1,\phi_0,\phi_1)$ be a term-algebra. To define
substitution algebras, we will consider
\begin{align*}
\jhom*
  {{{{\Gamma}{A}}{P}}{T}}
  {\ctxwk{T}{\jcomp{}{\epsilon_0}{P}}}
  {P}
  {\sigma_0}
  \\
\jfhom*
  {{{{\Gamma}{A}}{P}}{T}}
  {\ctxwk{T}{\jcomp{}{\epsilon_0}{P}}}
  {P}
  {\sigma_0}
  {\ctxwk{T}{\jcomp{}{\epsilon_0}{\jcomp{}{\epsilon_0}{P}}}}
  {\jcomp{}{\epsilon_0}{P}}
  {\sigma_1}
  \\
\jfhom*
  {{{{\Gamma}{A}}{P}}{T}}
  { \ctxext
      {\ctxwk{T}{\jcomp{}{\epsilon_0}{P}}}
      {\ctxwk{T}{\jcomp{}{\epsilon_0}{\jcomp{}{\epsilon_0}{P}}}}
    }
  {\ctxext{P}{\jcomp{}{\epsilon_0}{P}}}
  {\jvcomp{}{\sigma_0}{\sigma_1}}
  {\ctxwk{T}{\jcomp{}{\epsilon_0}{\jcomp{}{\epsilon_0}{T}}}}
  {\jcomp{}{\epsilon_0}{T}}
  {\sigma_2}
\end{align*}
for which the pair $(\sigma_0,\sigma_1)$ is required to be an extension
homomorphism and the triple $(\sigma_0,\sigma_1,\sigma_2)$ is required to
be an empty term-homomorphism.
